---
title: "Predicting Classe in Barbell Lifts"
author: "Bill Zichos"
date: "September 25, 2016"
output:
  html_document:
    keep_md: yes
---

I developed a model that with >99% accuracy, predicts the classe for different barbell lifts.  The best model used gradient boosting and 40 predictors including some features engineered from the date.  The rest of this document is organized in the following sections.

- Setup

- Analysis

- Model and Feature Selection

- Prediction

#Setup

```{r, message=FALSE, warning=FALSE}
library("caret")
library("Metrics")
library("e1071")
library("lubridate")

setwd("~/Coursera-Data-Science/08 Practical Machine Learning")

set.seed(10)
```

##Gather the data.

```{r}
train <- read.csv("pml-training.csv", stringsAsFactors = FALSE)
test <- read.csv("pml-testing.csv", stringsAsFactors = FALSE)

inSplit <- createDataPartition(train$classe, p = .7, list = FALSE)
```

##Prepare the data.

*X* is a unique identifier, so we want to make sure it is excluded from model fitting.

```{r}
exclude.predictors <- which(names(train)=="X")
```

Drop predictors that have more than 19000 NULLS.  This is way too sparse to add any value.

```{r}
toomanynulls <- which(sapply(train, function(x) {length(x[is.na(x)])}) > 19000)

train <- train[,-toomanynulls]
test <- test[,-toomanynulls]
```

*cvtd_timestamp* needs to be formatted as a datetime.

```{r}
train$cvtd_timestamp <- as.POSIXct(train$cvtd_timestamp, format = "%d/%m/%Y %R")
train$cvtd_timestamp.year <- year(train$cvtd_timestamp)
train$cvtd_timestamp.month <- month(train$cvtd_timestamp)
train$cvtd_timestamp.day <- day(train$cvtd_timestamp)
train$cvtd_timestamp.hour <- hour(train$cvtd_timestamp)
train$cvtd_timestamp.minute <- minute(train$cvtd_timestamp)
train$cvtd_timestamp <- NULL

test$cvtd_timestamp <- as.POSIXct(test$cvtd_timestamp, format = "%d/%m/%Y %R")
test$cvtd_timestamp.year <- year(test$cvtd_timestamp)
test$cvtd_timestamp.month <- month(test$cvtd_timestamp)
test$cvtd_timestamp.day <- day(test$cvtd_timestamp)
test$cvtd_timestamp.hour <- hour(test$cvtd_timestamp)
test$cvtd_timestamp.minute <- minute(test$cvtd_timestamp)
test$cvtd_timestamp <- NULL
```

Center and scale the numeric predictors.

```{r}
train$roll_belt <- scale(train$roll_belt, center = TRUE, scale = TRUE)
test$roll_belt <- scale(test$roll_belt, center = TRUE, scale = TRUE)
train$pitch_belt <- scale(train$pitch_belt, center = TRUE, scale = TRUE)
test$pitch_belt <- scale(test$pitch_belt, center = TRUE, scale = TRUE)
train$yaw_belt <- scale(train$yaw_belt, center = TRUE, scale = TRUE)
test$yaw_belt <- scale(test$yaw_belt, center = TRUE, scale = TRUE)
train$gyros_belt_x <- scale(train$gyros_belt_x, center = TRUE, scale = TRUE)
test$gyros_belt_x <- scale(test$gyros_belt_x, center = TRUE, scale = TRUE)
train$gyros_belt_y <- scale(train$gyros_belt_y, center = TRUE, scale = TRUE)
test$gyros_belt_y <- scale(test$gyros_belt_y, center = TRUE, scale = TRUE)
train$gyros_belt_z <- scale(train$gyros_belt_z, center = TRUE, scale = TRUE)
test$gyros_belt_z <- scale(test$gyros_belt_z, center = TRUE, scale = TRUE)
train$roll_arm <- scale(train$roll_arm, center = TRUE, scale = TRUE)
test$roll_arm <- scale(test$roll_arm, center = TRUE, scale = TRUE)
train$pitch_arm <- scale(train$pitch_arm, center = TRUE, scale = TRUE)
test$pitch_arm <- scale(test$pitch_arm, center = TRUE, scale = TRUE)
train$yaw_arm <- scale(train$yaw_arm, center = TRUE, scale = TRUE)
test$yaw_arm <- scale(test$yaw_arm, center = TRUE, scale = TRUE)
train$gyros_arm_x <- scale(train$gyros_arm_x, center = TRUE, scale = TRUE)
test$gyros_arm_x <- scale(test$gyros_arm_x, center = TRUE, scale = TRUE)
train$gyros_arm_y <- scale(train$gyros_arm_y, center = TRUE, scale = TRUE)
test$gyros_arm_y <- scale(test$gyros_arm_y, center = TRUE, scale = TRUE)
train$gyros_arm_z <- scale(train$gyros_arm_z, center = TRUE, scale = TRUE)
test$gyros_arm_z <- scale(test$gyros_arm_z, center = TRUE, scale = TRUE)
train$roll_dumbbell <- scale(train$roll_dumbbell, center = TRUE, scale = TRUE)
test$roll_dumbbell <- scale(test$roll_dumbbell, center = TRUE, scale = TRUE)
train$pitch_dumbbell <- scale(train$pitch_dumbbell, center = TRUE, scale = TRUE)
test$pitch_dumbbell <- scale(test$pitch_dumbbell, center = TRUE, scale = TRUE)
train$yaw_dumbbell <- scale(train$yaw_dumbbell, center = TRUE, scale = TRUE)
test$yaw_dumbbell <- scale(test$yaw_dumbbell, center = TRUE, scale = TRUE)
train$gyros_dumbbell_x <- scale(train$gyros_dumbbell_x, center = TRUE, scale = TRUE)
test$gyros_dumbbell_x <- scale(test$gyros_dumbbell_x, center = TRUE, scale = TRUE)
train$gyros_dumbbell_y <- scale(train$gyros_dumbbell_y, center = TRUE, scale = TRUE)
test$gyros_dumbbell_y <- scale(test$gyros_dumbbell_y, center = TRUE, scale = TRUE)
train$gyros_dumbbell_z <- scale(train$gyros_dumbbell_z, center = TRUE, scale = TRUE)
test$gyros_dumbbell_z <- scale(test$gyros_dumbbell_z, center = TRUE, scale = TRUE)
train$magnet_dumbbell_z <- scale(train$magnet_dumbbell_z, center = TRUE, scale = TRUE)
test$magnet_dumbbell_z <- scale(test$magnet_dumbbell_z, center = TRUE, scale = TRUE)
train$roll_forearm <- scale(train$roll_forearm, center = TRUE, scale = TRUE)
test$roll_forearm <- scale(test$roll_forearm, center = TRUE, scale = TRUE)
train$pitch_forearm <- scale(train$pitch_forearm, center = TRUE, scale = TRUE)
test$pitch_forearm <- scale(test$pitch_forearm, center = TRUE, scale = TRUE)
train$yaw_forearm <- scale(train$yaw_forearm, center = TRUE, scale = TRUE)
test$yaw_forearm <- scale(test$yaw_forearm, center = TRUE, scale = TRUE)
train$gyros_forearm_x <- scale(train$gyros_forearm_x, center = TRUE, scale = TRUE)
test$gyros_forearm_x <- scale(test$gyros_forearm_x, center = TRUE, scale = TRUE)
train$gyros_forearm_y <- scale(train$gyros_forearm_y, center = TRUE, scale = TRUE)
test$gyros_forearm_y <- scale(test$gyros_forearm_y, center = TRUE, scale = TRUE)
train$gyros_forearm_z <- scale(train$gyros_forearm_z, center = TRUE, scale = TRUE)
test$gyros_forearm_z <- scale(test$gyros_forearm_z, center = TRUE, scale = TRUE)
train$magnet_forearm_y <- scale(train$magnet_forearm_y, center = TRUE, scale = TRUE)
test$magnet_forearm_y <- scale(test$magnet_forearm_y, center = TRUE, scale = TRUE)
train$magnet_forearm_z <- scale(train$magnet_forearm_z, center = TRUE, scale = TRUE)
test$magnet_forearm_z <- scale(test$magnet_forearm_z, center = TRUE, scale = TRUE)
train$cvtd_timestamp.year <- scale(train$cvtd_timestamp.year, center = TRUE, scale = TRUE)
test$cvtd_timestamp.year <- scale(test$cvtd_timestamp.year, center = TRUE, scale = TRUE)
train$cvtd_timestamp.month <- scale(train$cvtd_timestamp.month, center = TRUE, scale = TRUE)
test$cvtd_timestamp.month <- scale(test$cvtd_timestamp.month, center = TRUE, scale = TRUE)

train$kurtosis_roll_belt <- scale(as.numeric(train$kurtosis_roll_belt), center = TRUE, scale = TRUE)
test$kurtosis_roll_belt <- scale(as.numeric(test$kurtosis_roll_belt), center = TRUE, scale = TRUE)

train$kurtosis_picth_belt <- scale(as.numeric(train$kurtosis_picth_belt), center = TRUE, scale = TRUE)
test$kurtosis_picth_belt <- scale(as.numeric(test$kurtosis_picth_belt), center = TRUE, scale = TRUE)

train$skewness_roll_belt <- scale(as.numeric(train$skewness_roll_belt), center = TRUE, scale = TRUE)
test$skewness_roll_belt <- scale(as.numeric(test$skewness_roll_belt), center = TRUE, scale = TRUE)

train$skewness_roll_belt.1 <- scale(as.numeric(train$skewness_roll_belt.1), center = TRUE, scale = TRUE)
test$skewness_roll_belt.1 <- scale(as.numeric(test$skewness_roll_belt.1), center = TRUE, scale = TRUE)

train$skewness_yaw_belt <- scale(as.numeric(train$skewness_yaw_belt), center = TRUE, scale = TRUE)
test$skewness_yaw_belt <- scale(as.numeric(test$skewness_yaw_belt), center = TRUE, scale = TRUE)
```

Center and scale the integer predictors.

```{r}
train$raw_timestamp_part_1 <- scale(train$raw_timestamp_part_1, center = TRUE, scale = TRUE)
test$raw_timestamp_part_1 <- scale(test$raw_timestamp_part_1, center = TRUE, scale = TRUE)
train$raw_timestamp_part_2 <- scale(train$raw_timestamp_part_2, center = TRUE, scale = TRUE)
test$raw_timestamp_part_2 <- scale(test$raw_timestamp_part_2, center = TRUE, scale = TRUE)
train$num_window <- scale(train$num_window, center = TRUE, scale = TRUE)
test$num_window <- scale(test$num_window, center = TRUE, scale = TRUE)
train$total_accel_belt <- scale(train$total_accel_belt, center = TRUE, scale = TRUE)
test$total_accel_belt <- scale(test$total_accel_belt, center = TRUE, scale = TRUE)
train$accel_belt_x <- scale(train$accel_belt_x, center = TRUE, scale = TRUE)
test$accel_belt_x <- scale(test$accel_belt_x, center = TRUE, scale = TRUE)
train$accel_belt_y <- scale(train$accel_belt_y, center = TRUE, scale = TRUE)
test$accel_belt_y <- scale(test$accel_belt_y, center = TRUE, scale = TRUE)
train$accel_belt_z <- scale(train$accel_belt_z, center = TRUE, scale = TRUE)
test$accel_belt_z <- scale(test$accel_belt_z, center = TRUE, scale = TRUE)
train$magnet_belt_x <- scale(train$magnet_belt_x, center = TRUE, scale = TRUE)
test$magnet_belt_x <- scale(test$magnet_belt_x, center = TRUE, scale = TRUE)
train$magnet_belt_y <- scale(train$magnet_belt_y, center = TRUE, scale = TRUE)
test$magnet_belt_y <- scale(test$magnet_belt_y, center = TRUE, scale = TRUE)
train$magnet_belt_z <- scale(train$magnet_belt_z, center = TRUE, scale = TRUE)
test$magnet_belt_z <- scale(test$magnet_belt_z, center = TRUE, scale = TRUE)
train$total_accel_arm <- scale(train$total_accel_arm, center = TRUE, scale = TRUE)
test$total_accel_arm <- scale(test$total_accel_arm, center = TRUE, scale = TRUE)
train$accel_arm_x <- scale(train$accel_arm_x, center = TRUE, scale = TRUE)
test$accel_arm_x <- scale(test$accel_arm_x, center = TRUE, scale = TRUE)
train$accel_arm_y <- scale(train$accel_arm_y, center = TRUE, scale = TRUE)
test$accel_arm_y <- scale(test$accel_arm_y, center = TRUE, scale = TRUE)
train$accel_arm_z <- scale(train$accel_arm_z, center = TRUE, scale = TRUE)
test$accel_arm_z <- scale(test$accel_arm_z, center = TRUE, scale = TRUE)
train$magnet_arm_x <- scale(train$magnet_arm_x, center = TRUE, scale = TRUE)
test$magnet_arm_x <- scale(test$magnet_arm_x, center = TRUE, scale = TRUE)
train$magnet_arm_y <- scale(train$magnet_arm_y, center = TRUE, scale = TRUE)
test$magnet_arm_y <- scale(test$magnet_arm_y, center = TRUE, scale = TRUE)
train$magnet_arm_z <- scale(train$magnet_arm_z, center = TRUE, scale = TRUE)
test$magnet_arm_z <- scale(test$magnet_arm_z, center = TRUE, scale = TRUE)
train$total_accel_dumbbell <- scale(train$total_accel_dumbbell, center = TRUE, scale = TRUE)
test$total_accel_dumbbell <- scale(test$total_accel_dumbbell, center = TRUE, scale = TRUE)
train$accel_dumbbell_x <- scale(train$accel_dumbbell_x, center = TRUE, scale = TRUE)
test$accel_dumbbell_x <- scale(test$accel_dumbbell_x, center = TRUE, scale = TRUE)
train$accel_dumbbell_y <- scale(train$accel_dumbbell_y, center = TRUE, scale = TRUE)
test$accel_dumbbell_y <- scale(test$accel_dumbbell_y, center = TRUE, scale = TRUE)
train$accel_dumbbell_z <- scale(train$accel_dumbbell_z, center = TRUE, scale = TRUE)
test$accel_dumbbell_z <- scale(test$accel_dumbbell_z, center = TRUE, scale = TRUE)
train$magnet_dumbbell_x <- scale(train$magnet_dumbbell_x, center = TRUE, scale = TRUE)
test$magnet_dumbbell_x <- scale(test$magnet_dumbbell_x, center = TRUE, scale = TRUE)
train$magnet_dumbbell_y <- scale(train$magnet_dumbbell_y, center = TRUE, scale = TRUE)
test$magnet_dumbbell_y <- scale(test$magnet_dumbbell_y, center = TRUE, scale = TRUE)
train$total_accel_forearm <- scale(train$total_accel_forearm, center = TRUE, scale = TRUE)
test$total_accel_forearm <- scale(test$total_accel_forearm, center = TRUE, scale = TRUE)
train$accel_forearm_x <- scale(train$accel_forearm_x, center = TRUE, scale = TRUE)
test$accel_forearm_x <- scale(test$accel_forearm_x, center = TRUE, scale = TRUE)
train$accel_forearm_y <- scale(train$accel_forearm_y, center = TRUE, scale = TRUE)
test$accel_forearm_y <- scale(test$accel_forearm_y, center = TRUE, scale = TRUE)
train$accel_forearm_z <- scale(train$accel_forearm_z, center = TRUE, scale = TRUE)
test$accel_forearm_z <- scale(test$accel_forearm_z, center = TRUE, scale = TRUE)
train$magnet_forearm_x <- scale(train$magnet_forearm_x, center = TRUE, scale = TRUE)
test$magnet_forearm_x <- scale(test$magnet_forearm_x, center = TRUE, scale = TRUE)
train$cvtd_timestamp.day <- scale(train$cvtd_timestamp.day, center = TRUE, scale = TRUE)
test$cvtd_timestamp.day <- scale(test$cvtd_timestamp.day, center = TRUE, scale = TRUE)
train$cvtd_timestamp.hour <- scale(train$cvtd_timestamp.hour, center = TRUE, scale = TRUE)
test$cvtd_timestamp.hour <- scale(test$cvtd_timestamp.hour, center = TRUE, scale = TRUE)
train$cvtd_timestamp.minute <- scale(train$cvtd_timestamp.minute, center = TRUE, scale = TRUE)
test$cvtd_timestamp.minute <- scale(test$cvtd_timestamp.minute, center = TRUE, scale = TRUE)
```

Format factors.

```{r}
train$user_name <- as.factor(train$user_name)
train$classe <- as.factor(train$classe)
train$new_window <- as.factor(train$new_window)
train$kurtosis_yaw_belt <- as.factor(train$kurtosis_yaw_belt)

test$user_name <- as.factor(test$user_name)
#test$classe <- as.factor(test$classe)
test$new_window <- as.factor(test$new_window)
test$kurtosis_yaw_belt <- as.factor(test$kurtosis_yaw_belt)
```

Impute missing values.

```{r}
train$kurtosis_roll_belt[is.na(train$kurtosis_roll_belt)] <- median(train$kurtosis_roll_belt[inSplit], na.rm = TRUE)
test$kurtosis_roll_belt[is.na(test$kurtosis_roll_belt)] <- median(test$kurtosis_roll_belt[inSplit], na.rm = TRUE)

train$kurtosis_picth_belt[is.na(train$kurtosis_picth_belt)] <- median(train$kurtosis_picth_belt[inSplit], na.rm = TRUE)
test$kurtosis_picth_belt[is.na(test$kurtosis_picth_belt)] <- median(test$kurtosis_picth_belt[inSplit], na.rm = TRUE)

train$skewness_roll_belt[is.na(train$skewness_roll_belt)] <- median(train$skewness_roll_belt[inSplit], na.rm = TRUE)
test$skewness_roll_belt[is.na(test$skewness_roll_belt)] <- median(test$skewness_roll_belt[inSplit], na.rm = TRUE)

train$skewness_roll_belt.1[is.na(train$skewness_roll_belt.1)] <- median(train$skewness_roll_belt.1[inSplit], na.rm = TRUE)
test$skewness_roll_belt.1[is.na(test$skewness_roll_belt.1)] <- median(test$skewness_roll_belt.1[inSplit], na.rm = TRUE)
```

Split the training data for cross validation.

```{r}
train1 <- train[inSplit, -exclude.predictors]
train2 <- train[-inSplit, -exclude.predictors]
```

Remove zero- and near zero-variance predictors.

```{r}
nzv <- nearZeroVar(train1)

train1 <- train1[, -nzv]
train2 <- train2[, -nzv]
```


#Analysis

```{r}
summary(train$classe)
```

```{r, warning=FALSE, message=FALSE}
fit.rpart <- train(classe~., data = train1, method = "rpart")
fit.gbm <- train(classe ~ ., data = train1, method = "gbm")
#fit.svm <- svm(classe ~ ., train1)
#fit.rf <- train(classe ~ ., data = train1, method = "rf")
fit.lda <- train(classe ~ ., data = train1, method = "lda")
```

```{r}
pred.rpart <- predict(fit.rpart, train2)
pred.gbm <- predict(fit.gbm, train2)
#pred.svm <- predict(fit.svm, train2)
#pred.rf <- predict(fit.rf, train2)
pred.lda <- predict(fit.lda, train2)
```

```{r}
confusionMatrix(pred.rpart, train2$classe)
confusionMatrix(pred.gbm, train2$classe)
#confusionMatrix(pred.svm, train2$classe)
#confusionMatrix(pred.rf, train2$classe)
confusionMatrix(pred.lda, train2$classe)
```

#Model and Feature Selection

Recursive Partitioning/Decision Trees Accuracy: **0.5998**

Gradient Boosting (GBM) Accuracy: **0.9954**

Random Forest Accuracy: **NA**

Linear Discriminant Analysis (LDA) Accuracy: **0.8413**

At 99.54% Accuracy, we will use the GBM model for predicting new observations.

The best predictors for this dataset are below.

```{r}
predictors(fit.gbm)
```

#Prediction

```{r}
test$prediction <- predict(fit.gbm, test)
test[,which(names(test) %in% c("problem_id", "prediction"))]
```

#Appendix

```{r}
#str(train)

summary(train)
```